Modelling the gas kinematics of an atypical Lyman alpha emitting compact dwarf galaxy

### 2017-06-26 paper/paper.tex
small changes

\begin{abstract}
  The conditions in star-forming Compact Dwarf Galaxies (CDGs) 
  place them as the best systems to test models on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya\ line emission that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these kinematic conditions.
The multiphase model requires a clump velocity dispersion of
at least \sigmaclump\ with outflows of \inftyclump\ at most, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
This shows the importance of considering multiphase
and rotation kinematics among the possible conditions shaping the
\lya\ spectra.
However, we consider that the multiphase model has more chances to
agree with \tol's reality. 
The clump velocity dispersion implies a dynamical mass of at least $2\times
10^{9}$ M$_{\odot}$, ten times its baryonic mass, which can be
explained if \tol\ is hosted by a dark matter halo of at least
$10^{11}$ M$_{\odot}$; conditions similar to those of Dwarf Elliptical
Galaxies. 
If future kinematic maps of \tol\ confirm the velocities found by the
multiphase model, it could provide a new perspective to study
evolutionary processes linking star-forming with quiescent phases in
primordial galaxy evolution. 
\end{abstract}

@@

The motivation for the rotation without outflows (as presented in \cite{GaravitoCamargo2014})is that
dwarf galaxies show coherent rotation features
\citep{2009A&A...493..871S} and it is expected that some of them have
high neutral gas contents with long quiescent phases without high gas
outflows triggered by supernova activity
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}.

@@

\begin{table}
\begin{center}
\begin{tabular}{lc}\hline
$\alpha$(2000)$^{a}$ & 12h17min17.1s\\
$\delta$(2000)$^{b}$ & -28d02m32s\\
$l$, $b$ (deg) & 294, 34\\
$m_V$ & 17.5\\
  M$_V$ & -17.6\\ 
$v$(km s$^{-1}$) & $7785\pm 50$\\
$z$ & $0.0259\pm0.0002$ \\
\lya\ (erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$)& $8.1\times 10^{-14}$ \\
\lya\ Equivalent Width (\AA) & $70$\\
$21$cm (Jy km s$^{-1}$)& $<0.10$ \\\hline
\end{tabular}
\end{center}
\caption{Basic observational characteristics of \tol\ 
  \citep{Thuan97}\label{obstable}} 
\end{table}

@@

The observed flux for the \lya\ line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \citep{Thuan97}
and a Equivalent Width of $70$\AA\ and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
\citep{Izotov04} which gives a \lya/H$\beta$ flux ratio of
4.9$\pm$0.1.
The \lya\ flux values correspond to a luminosity of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$, which in turn
translates  into a lower bound for the star formation rate of $2.0$
M$_{\odot}$ yr$^{-1}$ after using a standard conversion factor between
luminosity and star formation rate of $9.1\times 10^{-43}$
$L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$ \citep{Kennicutt98} without any
correction by extinction.  
Comparing the \lya/H$\beta$ ratio with the theoretical
expectation from case B recombination of $23.3$ \citep{Hummer1987} one
can estimate an escape fraction of $20$\% for \lya\ radiation.

The bolometric UV luminosity is $9.43\pm1.94 \times 10^{8}$
L$_{\odot}$ as measured by GALEX, without any correction by extinction
and following the empirical relation by \cite{Kennicutt98}, this
corresponds to a star formation rate of $0.35\pm 0.05$ M$_{\odot}$
yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.  
Its metallicity is $\sim Z_{\odot}/24$ as derived from
optical spectroscopy \citep{Izotov04}. 
% using http://tomdwelly.com/tools_fluxtolum.php
%

@@

There is an upper limit for the  21 cm line integrated flux of $<0.10$
Jy km s$^{-1}$  which translates into a upper limit for the HI mass of
$M<2.65\times 10^{8}$ M$_{\odot}$ \citep{pustilnikmartin07}.

\begin{figure*}
\begin{center}
\includegraphics[width=0.95\textwidth]{tree_result.pdf}
\caption{Classification tree example. 
The aim is to find the parameters that separate two
classes: low $\chi^2/d.o.f. < 35$ and high $\chi^2/d.o.f>35$. 
Each box describes the condition over the parameter of interest, the
level of sample impurity, the number of samples and the value of the
number of samples in each classes.  
In this example we start with $28$ models in the first class and
$1962$ models in the second class; the best way to increase the probability to find a
a result with low $\chi^2/d.o.f.$ ($13$ models) is having $v_{\infty,{\rm cl}}<157.0$ \kms,
$\sigma_{{\rm cl}}>55.6$ \kms and $P_{\rm cl}>0.683$.
This classification example shows that all other parameters are also
important but these three are the most relevant. 
\label{fig:tree}}
\end{center}
\end{figure*}

@@

We cannot discard that another region of parameter space
could also work.
It might be interesting to explore in the
future with a higher sampling of the parameter space once new 
observations  clarify the kinematic nature of \tol \footnote{In
  Section \S \ref{sec:conclusions} we describe  in more detail what
  kind of observations could be performed in the near future to
  improve our understanding of \tol's kinematics.} .

Considering this sparse sampling of parameter space and the results
from the random forest classifier, it is reasonable to consider that
the best values for $\sigma_{\rm cl}$, $v_{\infty {\rm cl}}$, $P_{\rm
  cl}$ are actually lower/upper/lower bounds, respectively.

@@

Which model has more
chances to be correct?  
We discard the rotation model by considering that the rotational
velocity of $\sim350$ \kms\ is too high for such a low
mass system. 
\tol's infered rotational velocity is larger than the
Milky Way's ($\sim200$\kms) while its baryonic mass is one order
of magnitude smaller. 
Furthermore, \tol's structural properties seem to be similar to known dwarf
elliptical galaxies. 
The sample of 7 dwarf ellipticals (dE) observed by
\cite{2003AJ....126.1794G} show similar properties as \tol. 
For instance, the dE galaxies IC 3344 and IC 794 have similar absolute $M_V$
magnitudes (-17.10 and -17.19) and sizes (0.90 kpc and 0.82 kpc) as
\tol. 
These dE galaxies have velocity dispersions of $31.1$ \kms and
$44.6$ \kms, respectively,  and their rotational velocities are lower
than $1$ \kms.  
This leads us to prefer the multiphase model as it provides at least
some consistency with kinematic properties of dE galaxies.

@@

\cite{2011ApJ...726..108T} showed that  that such dynamical
mass can be explained if a dwarf galaxy is hosted by a dark matter halo of at
least $10^{11}$ M$_{\odot}$ in mass. 
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the \emph{median} mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P}.

These results would imply that \tol\ is part of the dE
class. 
However, there are two puzzles that are opened up by this hypothesis. 
First, dwarf ellipticals are found either in clusters or groups of
galaxies; while \tol\ does not belong either to a group or a cluster
\citep{2001AJ....121..169F,}. 
The second problem is that, unlike dE, \tol\ hosts a bright and compact
startburst; imaging and spectroscopy studies suggest that the
startburst started less than 4 Myr ago, while the low surface
brightness regins are consistent with an age below one Gyr
\citep{2001AJ....121..169F}.

@@

In this paper we presented two kinematic models that independently
reproduce the so far unexplained observational features of \tol's
\lya\ emission line.  
One model is based on a multiphase ISM with random clump motions and
the other on gas bulk solid body rotation.

It is the first time that an observed \lya\ profile can be fully
reproduced by either of these two kinematic conditions.
We give preference to the multiphase model as it produces kinematic
conditions that are similar to those in dwarf ellipticals, while the
rotational model produces rotational velocities to high for a dwarf
galaxy.

However, \tol's environment and star forming activity are radically
different from dwarf ellipticals.
We suggest that additional integral field unit observations are the
only way to be certain about the detailed kinematic structure in \tol.

All in all, the mere existence of a strong LAE galaxy with a broad,
symmetric line is interesting. 
\tol's line shape is different to others and reveals an unexpected
kinematic structure.
Our findings highlight the importance of including multiphase and/or
rotation conditions as kinematic features to model the \lya\ line.

A definitive confirmation of our kinematic findings has
to be provided by new observational tests with integral field unit
spectroscopy. 
If our results are confirmed, this would open up a new line of more detailed
enquiries to understand how a compact dwarf galaxy can have the
structural characteristics similar to a dwarf elliptical while being
at the same time isolated and sustaining large starburst.
Such and investigation could provide an evolutionary link between star-forming
and quiescent stages of dwarf galaxy evolution.

\section*{Acknowledgments}
We thank J. M. Mas-Hesse for providing us with \tol's observational
\lya\ data \citep{mashesse03} in electronic format. This data was
used to prepare Figure 1.
J. E. F-R. ackwnoledges the support of J. L. Guerra in the last
writing stages of this paper.

>>>>

\begin{abstract}
  The conditions in star-forming Compact Dwarf Galaxies (CDGs) 
  place them as the best systems to test models on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya\ line emission that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these kinematic conditions.
The multiphase model requires a clump velocity dispersion of
at least \sigmaclump\ with outflows of \inftyclump\ at most, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
This shows the importance of considering multiphase
and rotation kinematics among the possible conditions shaping the
\lya\ spectra.
However, we consider that the multiphase model has more chances to
agree with \tol's reality. 
The clump velocity dispersion implies a dynamical mass of at least $2\times
10^{9}$ M$_{\odot}$, ten times its baryonic mass, which can be
explained if \tol\ is hosted by a dark matter halo of at least
$10^{11}$ M$_{\odot}$; conditions similar to those of dwarf elliptical
galaxies. 
If future kinematic maps of \tol\ confirm the velocities found by the
multiphase model, it could provide a new perspective to study
evolutionary processes linking star-forming with quiescent phases in
primordial galaxy evolution. 
\end{abstract}

@@

The motivation for the rotation without outflows (as presented in
\cite{GaravitoCamargo2014}) is that dwarf galaxies show coherent
rotation features  \citep{2009A&A...493..871S} and it is expected that
some of them have high neutral gas contents with long quiescent phases
without high gas outflows triggered by supernova activity
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}.

@@

\begin{table}
\begin{center}
\begin{tabular}{lc}\hline
$\alpha$(2000)$^{a}$ & 12h17min17.1s\\
$\delta$(2000)$^{b}$ & -28d02m32s\\
$l$, $b$ (deg) & 294, 34\\
$m_V$ & 17.5\\
  M$_V$ & -17.6\\ 
$z$ & $0.0259\pm0.0002$ \\\hline
\end{tabular}
\end{center}
\caption{Basic observational characteristics of \tol\ 
  \citep{Thuan97}\label{obstable}} 
\end{table}

@@

The observed flux for the \lya\ line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \citep{Thuan97}.
The \lya\ Equivalent Width is $70$\AA\ and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
which gives a \lya/H$\beta$ flux ratio of
4.9$\pm$0.1 \citep{Izotov04}.
The \lya\ flux values correspond to a luminosity of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$, which in turn
translates  into a lower bound for the star formation rate of $2.0$
M$_{\odot}$ yr$^{-1}$ after using a standard conversion factor between
luminosity and star formation rate of $9.1\times 10^{-43}$
$L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$ 
without any
correction by extinction
\citep{Kennicutt98}.
Comparing the \lya/H$\beta$ ratio with the theoretical
expectation from case B recombination of $23.3$ \citep{Hummer1987} one
can estimate an escape fraction of $20$\% for \lya\ radiation.

The bolometric UV luminosity is $9.43\pm1.94 \times 10^{8}$
L$_{\odot}$ as measured by GALEX. Without any correction by extinction
and following the empirical relation by \cite{Kennicutt98}, this
corresponds to a star formation rate of $0.35\pm 0.05$ M$_{\odot}$
yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.  
Its metallicity is $\sim Z_{\odot}/24$ as derived from
optical spectroscopy \citep{Izotov04}. 
% using http://tomdwelly.com/tools_fluxtolum.php
%

@@

There is an upper limit for the  $21$ cm line integrated flux of $<0.10$
Jy km s$^{-1}$  which translates into a upper limit for the HI mass of
$M<2.65\times 10^{8}$ M$_{\odot}$ \citep{pustilnikmartin07}.

\begin{figure*}
\begin{center}
\includegraphics[width=0.95\textwidth]{tree_result.pdf}
\caption{Classification tree example. 
The aim is to find the parameters that separate two
classes: low $\chi^2/d.o.f. < 35$ and high $\chi^2/d.o.f>35$. 
Each box describes the condition over the parameter of interest, the
level of sample impurity, the number of samples and the value of the
number of samples in each classes.  
In this example we randomly sample $80\%$ of the full set of models to
start with $28$ models in the first class and
$1962$ models in the second class; the best way to increase the
probability to find a  a result with low $\chi^2/d.o.f.$ ($13$ models) is having $v_{\infty,{\rm cl}}<157.0$ \kms,
$\sigma_{{\rm cl}}>55.6$ \kms and $P_{\rm cl}>0.683$.
The final classification is not perfect, showing that all other
parameters are also important but these three are the most relevant.  
\label{fig:tree}}
\end{center}
\end{figure*}

@@

We cannot discard that another region of parameter space
could also work.
It might be interesting to explore in the
future with a higher sampling of the parameter space once new 
observations  clarify the kinematic nature of \tol \footnote{In
  Section \S \ref{sec:conclusions} we describe  in more detail what
  kind of observations could be performed in the near future to
  improve our understanding of \tol's kinematics.} . 
Considering the sparse sampling of parameter space and the results
from the random forest classifier, it is safer to consider that
the best values for $\sigma_{\rm cl}$, $v_{\infty {\rm cl}}$, $P_{\rm
  cl}$ are actually lower/upper/lower bounds, respectively.

@@

Which model has more
chances to be correct?  
We discard the rotation model by considering that the rotational
velocity of $\sim350$ \kms\ is too high for such a low
mass system. 
\tol's infered rotational velocity is larger than the
Milky Way's ($\sim200$\kms) while its baryonic mass is one order
of magnitude smaller. 
Furthermore, \tol's structural properties seem to be similar to known dwarf
elliptical galaxies. 
The sample of 7 dwarf ellipticals (dE) observed by
\cite{2003AJ....126.1794G} show similar properties as \tol. 
For instance, the dE galaxies IC 3344 and IC 794 have similar absolute $M_V$
magnitudes (-17.10 and -17.19) and sizes (0.90 kpc and 0.82 kpc) as
\tol. 
These dE galaxies have velocity dispersions of $31.1$ \kms and
$44.6$ \kms, respectively,  and their rotational velocities are lower
than $1$ \kms.  
This leads us to prefer the multiphase model as it provides 
a broad consistency with kinematic properties of dE galaxies.

@@

\cite{2011ApJ...726..108T} showed that  that such dynamical
mass can be explained if a dwarf galaxy is hosted by a dark matter halo of at
least $10^{11}$ M$_{\odot}$ in mass. 
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the median mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P}.

These results would imply that \tol\ is part of the dE
class. 
However, there are two puzzles that are opened up by this hypothesis. 
First, dE galaxies are found either in clusters or groups of
galaxies; while \tol\ does not belong either to a group or a cluster
\citep{2001AJ....121..169F}. 
The second problem is that, unlike a dE, \tol\ hosts a bright and compact
startburst; imaging and spectroscopy studies suggest that the
startburst started less than 4 Myr ago, while the low surface
brightness regins are consistent with an age below one Gyr
\citep{2001AJ....121..169F}.

@@

In this paper we presented two kinematic models that independently
reproduce the so far unexplained observational features of \tol's
\lya\ emission line.  
One model is based on a multiphase ISM with random clump motions and
the other on gas bulk solid body rotation. 
It is the first time that an observed \lya\ profile can be fully
reproduced by either of these two kinematic conditions.
Our findings highlight the importance of including multiphase and/or
rotation conditions as kinematic features to model the \lya\ line.

We prefer the multiphase model becasue it has kinematic
conditions similar to those in dwarf ellipticals, while the
rotational model produces rotational velocities to high for a dwarf
galaxy.

However, \tol's environment and star forming activity are radically
different from those of dwarf ellipticals.
We suggest that additional integral field unit observations are the
only way to be certain about the detailed kinematic structure in \tol.

All in all, the mere existence of a strong LAE galaxy with a broad,
symmetric line is interesting. 
\tol's line shape is different to others and reveals an unexpected
kinematic structure.
A definitive confirmation of our kinematic findings has
to be provided by new observational tests with integral field unit
spectroscopy. 
If our results are confirmed, this would give robustness to our models
as tools to study the evolutionary link between star-forming and
quiescent stages of dwarf galaxy evolution.

\section*{Acknowledgments}
We thank the referee for detailed comments and suggestions that
improved this paper. 
We thank J. M. Mas-Hesse for providing us with \tol's observational
\lya\ data \citep{mashesse03} in electronic format. This data was
used to prepare Figure 1.
J. E. F-R. ackwnoledges the support of J. L. Guerra in the last
writing stages of this paper.
